---
title: "EDA HINTS DATA"
format:
  html: 
    embed-resources: true
---

```{r, message = FALSE, warning = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(gtsummary)

```


# Introduction

In the digital age, where information is readily available at our fingertips, understanding how people navigate and trust health information is critical. With countless sources such as healthcare professionals, government agencies, social media, and online forums, it is essential to explore what the United States population believes about the reliability and accessibility of health information. To better understand this phenomenon, data was gathered from an online forum called Reddit. The Reddit dataset comprises textual data that offers valuable insights into the information shared online about health, particularly topics related to cancer. Using NLP methods, we can explore the most common topics and also the sentiment toward these topics. In addition to the textual data, we will also use the Health Information National Trends Survey (HINTS), which is administered by National Cancer Institute. In this survey, people were asked directly about what they think about health information. Using both the Reddit data and HINTS, we seek to understand trust, frustrations, and overall sentiment about healthcare and more specifically, cancer information.  

The Health Information National Trends Survey (HINTS) is a national survey administered by the National Cancer Institute. The survey is completed by civilians in the United States. The information gathered from the survey seeks to gain an understanding of what people think about the accessibility of health information and the perceptions of what people think about the health information. The HINTS survey includes 527 questions. To focus on sentiment toward healthcare and cancer, we selected the questions in Table 1. In Table 1, participants were asked various questions about where they access health information, do they trust the provided health information, and if they feel frustrated about the information. 

**Table 1**

| **Variable**                | **Description**                                                                                           |
|-----------------------------|-----------------------------------------------------------------------------------------------------------|
| SeekCancerInfo              | Have you ever looked for information about cancer from any source?                                       |
| CancerFrustrated            | Based on the results of your most recent search for information about cancer, how much do you agree or disagree: You felt frustrated during your search for the information. |
| CancerTrustDoctor           | In general, how much would you trust information about cancer from a doctor?                             |
| CancerTrustFamily           | In general, how much would you trust information about cancer from family or friends?                    |
| CancerTrustGov              | In general, how much would you trust information about cancer from government health agencies?           |
| CancerTrustCharities        | In general, how much would you trust information about cancer from charitable organizations?             |
| CancerTrustReligiousOrgs    | In general, how much would you trust information about cancer from religious organizations and leaders?   |
| CancerTrustScientists       | In general, how much would you trust information about cancer from scientists?                           |
| Electronic2_HealthInfo      | In the past 12 months have you used the Internet to look for health or medical information?              |
| MisleadingHealthInfo        | How much of the health information that you see on social media do you think is false or misleading?     |
| TrustHCSystem               | How much do you trust the health care system (for example, hospitals, pharmacies, and other organizations involved in health care)? |


```{r}

load('../data/HINTS6_R_20240524/hints6_public.rda')
hints <- as.data.frame(public)

#print(colnames(hints))

columns <- c("HHID", "updatedate", "SeekCancerInfo", "CancerFrustrated", "CancerTrustDoctor", "CancerTrustFamily", "CancerTrustGov", "CancerTrustCharities", "CancerTrustReligiousOrgs", "CancerTrustScientists", "Electronic2_HealthInfo", "MisleadingHealthInfo", "TrustHCSystem")

hints_select <- hints %>% select(all_of(columns))
#hints_select$updatedate <- hints_select$updatedate / 1000
#hints_select$updatedate <- as_datetime(hints_select$updatedate)

head(hints_select)

```

# Survey Responses

In the bar graphs, a first look at the data provides an general overview of the responses to the questions. These plots show how much participants agree or disagree to each question. For example, many participants can trust doctors a lot and less than family members.  

```{r}

plot_data <- hints_select %>%
  select(-HHID, -updatedate) %>%  # Exclude the first two columns
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") %>%
  count(Variable, Value)

print(unique(plot_data$Value))

values <- c("Strongly agree", "Somewhat agree", "Somewhat disagree", "Strongly disagree", "A lot", "Some", 
            "A little", "Not at all", "Yes", "No", "None", "I do not use social media", "Very", "Somewhat")

plot_data_filtered <- plot_data %>% filter(Value %in% values) 
plot_data_filtered$Value <- factor(plot_data_filtered$Value, levels = sort(unique(plot_data_filtered$Value)))

columns_1 <- c("SeekCancerInfo", "CancerFrustrated", "CancerTrustDoctor", "CancerTrustFamily")
plot_data_filtered_1 <- plot_data_filtered %>% filter(Variable %in% columns_1)
columns_2 <- c("CancerTrustGov", "CancerTrustCharities", "CancerTrustReligiousOrgs", "CancerTrustScientists")
plot_data_filtered_2 <- plot_data_filtered %>% filter(Variable %in% columns_2)
columns_3 <- c("Electronic2_HealthInfo", "MisleadingHealthInfo", "TrustHCSystem")
plot_data_filtered_3 <- plot_data_filtered %>% filter(Variable %in% columns_3) 

p <- ggplot(plot_data_filtered_1, aes(x = Value, y = n, fill = Variable)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ Variable, scales = "free_x") +  # Separate plots for each column
  theme_minimal() +
  labs(
    title = "HINTS Survey Responses",
    x = "Responses",
    y = "Count",
    fill = "Question"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

p

png("../data/HINTS6_R_20240524/HINTS_plot1.png", width = 800, height = 600)
print(p)  
dev.off()
```

```{r}
p <- ggplot(plot_data_filtered_2, aes(x = Value, y = n, fill = Variable)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ Variable, scales = "free_x") +  
  theme_minimal() +
  labs(
    title = "HINTS Survey Responses",
    x = "Responses",
    y = "Count",
    fill = "Question"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

p

png("../data/HINTS6_R_20240524/HINTS_plot2.png", width = 800, height = 600)
print(p)  
dev.off()
```

```{r}
p <- ggplot(plot_data_filtered_3, aes(x = Value, y = n, fill = Variable)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ Variable, scales = "free_x", nrow=2) +
  theme_minimal() +
  labs(
    title = "HINTS Survey Responses",
    x = "Responses",
    y = "Count",
    fill = "Question"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

p

png("../data/HINTS6_R_20240524/HINTS_plot3.png", width = 800, height = 600)
print(p) 
dev.off()
```

# Summary Statistics of the Responses

The responses of the questions can be coded into a scale from 0:3. For example, "Not at all" is coded as 0, "A little" is coded as 1, "Some" is coded as 2, and "A lot" is coded as 3. After coding these responses, the mean, median, and mode is calculated to highlight overall trends in the data.  

```{r}
print(unique(hints_select$MisleadingHealthInfo))
```

```{r}
print(colnames(hints_select))

hints_select_coded <- hints_select %>%
  mutate(CancerFrustrated = as.numeric(case_when(
    CancerFrustrated == "Strongly disagree" ~ "0",
    CancerFrustrated == "Somewhat disagree" ~ "1",
    CancerFrustrated == "Somewhat agree" ~ "2",
    CancerFrustrated == "Strongly agree" ~ "3",
    TRUE ~ 	NA 
  )))

hints_select_coded <- hints_select_coded %>%
  mutate(CancerTrustDoctor = as.numeric(case_when(
    CancerTrustDoctor == "Not at all" ~ "0",
    CancerTrustDoctor == "A little" ~ "1",
    CancerTrustDoctor == "Some" ~ "2",
    CancerTrustDoctor == "A lot" ~ "3",
    TRUE ~ 	NA 
  )))

hints_select_coded <- hints_select_coded %>%
  mutate(CancerTrustFamily = as.numeric(case_when(
    CancerTrustFamily == "None" ~ "0",
    CancerTrustFamily == "A little" ~ "1",
    CancerTrustFamily == "Some" ~ "2",
    CancerTrustFamily == "A lot" ~ "3",
    TRUE ~ NA 
  )))

hints_select_coded <- hints_select_coded %>%
  mutate(SeekCancerInfo = as.numeric(case_when(
    SeekCancerInfo == "Yes" ~ "1",
    SeekCancerInfo == "No" ~ "0",
    TRUE ~ NA 
  )))

hints_select_coded <- hints_select_coded %>%
  mutate(CancerTrustCharities = as.numeric(case_when(
    CancerTrustCharities == "Not at all" ~ "0",
    CancerTrustCharities == "A little" ~ "1",
    CancerTrustCharities == "Some" ~ "2",
    CancerTrustCharities == "A lot" ~ "3",
    TRUE ~ NA 
  )))

hints_select_coded <- hints_select_coded %>%
  mutate(CancerTrustGov = as.numeric(case_when(
    CancerTrustGov == "Not at all" ~ "0",
    CancerTrustGov == "A little" ~ "1",
    CancerTrustGov == "Some" ~ "2",
    CancerTrustGov == "A lot" ~ "3",
    TRUE ~ NA
  )))

hints_select_coded <- hints_select_coded %>%
  mutate(CancerTrustReligiousOrgs = as.numeric(case_when(
    CancerTrustReligiousOrgs == "Not at all" ~ "0",
    CancerTrustReligiousOrgs == "A little" ~ "1",
    CancerTrustReligiousOrgs == "Some" ~ "2",
    CancerTrustReligiousOrgs == "A lot" ~ "3",
    TRUE ~ NA 
  )))

hints_select_coded <- hints_select_coded %>%
  mutate(CancerTrustScientists = as.numeric(case_when(
    CancerTrustScientists == "Not at all" ~ "0",
    CancerTrustScientists == "A little" ~ "1",
    CancerTrustScientists == "Some" ~ "2",
    CancerTrustScientists == "A lot" ~ "3",
    TRUE ~ NA 
  )))

hints_select_coded <- hints_select_coded %>%
  mutate(Electronic2_HealthInfo = as.numeric(case_when(
    Electronic2_HealthInfo == "Yes" ~ "1",
    Electronic2_HealthInfo == "No" ~ "0",
    TRUE ~ NA 
  )))

hints_select_coded <- hints_select_coded %>%
  mutate(MisleadingHealthInfo = as.numeric(case_when(
    MisleadingHealthInfo == "None" ~ "0",
    MisleadingHealthInfo == "I do not use social media" ~ "0",
    MisleadingHealthInfo == "A little" ~ "1",
    MisleadingHealthInfo == "Some" ~ "2",
    MisleadingHealthInfo == "A lot" ~ "3",
    TRUE ~ NA 
  )))

hints_select_coded <- hints_select_coded %>%
  mutate(TrustHCSystem = as.numeric(case_when(
    TrustHCSystem == "Not at all" ~ "0",
    TrustHCSystem == "A little" ~ "1",
    TrustHCSystem == "Somewhat" ~ "2",
    TrustHCSystem == "Very" ~ "3",
    TRUE ~ NA 
  )))


print(head(hints_select_coded))
```

In the summary table below, the mean for trusting a doctor is higher than trusting the government. Given this information, we will also look at the Reddit dataset to see the level of trust users have when they mention the government in their comments versus doctors. In addition, the people who felt frustrated about the information they received about cancer is approximately 1.105. In the Reddit dataset, we also look for an equivalent using textual data by looking at positive/negative and emotion sentiment analysis on comments that include the word "cancer".  

```{r}

print(summary(hints_select_coded))


```

```{r}

hints_select_coded_clean <- drop_na(hints_select_coded)

table1 <- hints_select_coded_clean %>%
  select(-HHID, -updatedate) %>%
  tbl_summary(
    statistic = all_continuous() ~ "{mean} ± {sd}",  
  )

table1

```


The box plot provides a visualization of the median, mode, and outliers in the dataset.  

```{r}


boxplot_data <- hints_select_coded %>%
  select(-HHID, -updatedate)


boxplot_data_long <- boxplot_data %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value")


ggplot(boxplot_data_long, aes(x = Variable, y = Value)) +
  geom_boxplot(outlier.colour = "red", outlier.size = 1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  
  labs(
    title = "HINTS Boxplot",
    x = "Variables",
    y = "Values"
  ) + coord_cartesian(ylim = c(-1, 4)) 

```


# Relationship between different survey questions.

Using a correlation plot, the relationships are evaluated between the different survey questions. In the correlation plot below, we see an negative correlation between trust for doctors versus trusting the government, scientists, and the healthcare system. Further statistical testing can be performed to better understand this initial evaluation.

```{r}

library(corrplot)

cor_matrix <- drop_na(hints_select_coded)

# correlation matrix
cor_matrix <- cor(cor_matrix[, sapply(cor_matrix, is.numeric)], use = "complete.obs")

# correlation plot
corrplot(cor_matrix, method = "circle", type = "upper", tl.col = "black", tl.srt = 45, 
         addCoef.col = "black", number.cex = 0.7, diag = FALSE)
```

# Further Directions and Conclusion

The HINTS dataset provides insight into the perceptions of healthcare and cancer information. The trends in this dataset will be repeated in the Reddit dataset. Using the Reddit dataset, we will explore sentiments, such as positive/negative, frustrations, and trust. We will also look at word frequency count to review, which topics Reddit users commonly comment about.  






